Oregon: Zip-Level Wildfire Smoke and Health Outcomes (open in script)

### 2017-11-03 r_scripts/analysis/county_timeseries.Rmd

```{r }
library(tidyverse)
library(readr)
library(lme4)
library(broom)
library(htmlTable)

@@

# analysis dataframe
or_may_sep_df <- or_ts_df %>% 
  # create day of the week variable
  mutate(day = as.factor(weekdays(date))) %>% 
  # set missing outcome values to 0 since missing indicates no ER or urgent care 
  # visits on that date (which is reasonable in sparsely populated counties)
  mutate_each(funs(wo_miss = ifelse(is.na(.), 0, .)), n_obs:broken_arm_n) %>% 
  # create a binary smoke indicator variable based smk variables >5, >10, and >15 units 
  mutate(wrf_smk5 = ifelse(wrf_smk_pm >= 5, 1, 0), 
         wrf_smk10 = ifelse(wrf_smk_pm >= 10, 1, 0),
         wrf_smk15 = ifelse(wrf_smk_pm >= 15, 1, 0), 
         geo_smk5 = ifelse(geo_smk_pm >= 5, 1, 0),
         geo_smk10 = ifelse(geo_smk_pm >= 10, 1, 0),
         geo_smk15 = ifelse(geo_smk_pm >= 15, 1, 0),
         krig_smk5 = ifelse(krig_smk_pm >= 5, 1, 0),
         krig_smk10 = ifelse(krig_smk_pm >= 10, 1, 0),
         krig_smk15 = ifelse(krig_smk_pm >= 15, 1, 0),
         season = ifelse(date >= "2013-05-01" & date <= "2013-06-21", "spring", 
                  ifelse(date >= "2013-06-22" & date <= "2013-09-22", "summer",
                  ifelse(date >= "2013-09-23" & date <= "2013-12-21", "fall",
                                "other")))) %>% 
  # rescale pm variables to 10 units
  mutate_each(funs(unit10 = ./10), wrf_pm:krig_smk_pm)
  # select(1:2, 31:33, 36, 34:35, 38:44, 28:29, 30, 55, 58, 
  #        65, 64, 63) # subset and order variables used in analysis

@@

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('Geo-Weighted Smoke', 'Kriging Smoke', 'WRF-Chem Smoke')

@@

#glimpse(wash_jul_oct_df)
# finding columns to retain
#which(colnames(wash_jul_oct_df)=='geo_smk_pm_unit10')
#summary(wash_jul_oct_df)
# loop through outcomes
for(i in 6:18){
  i = 8
  # empty matrix for table
  point_estimates <- matrix(nrow = 3, ncol = 5, byrow = T)

@@

  # loop through pm estimation methods
  for(j in 25:27){
    j = 25
    var_name <- method_list[j-22]

@@

    # populate matrix
    row_n <- j-22
    # method
    point_estimates[row_n, 2] <- var_name
    # rate ratio
    point_estimates[row_n, 3] <- round(exp(model_df[2,2]), 3)
    # 95% lower bound
    point_estimates[row_n, 4] <- round(exp((model_df[2,2]) -
                                             1.96*(model_df[2,3])), 3)
    # 95% upper bound
    point_estimates[row_n, 5] <- round(exp((model_df[2,2]) +
                                             1.96*(model_df[2,3])), 3)

@@

  # combine previous values in dataframe that has all outcome/methods comparisons
  datalist[[i]] <- point_est_df

@@

>>>>

###  County-Level Time-Series for the Fire Season
This time series analysis is for the Oregon county-level all emergency visit in All Payer All Claims data set.

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(lme4)
library(broom)
library(htmlTable)
library(knitr)
```

First filter the date to only 2013 May to 2013 September.
Rescale the pm variables to 10 units.

```{r, message=FALSE, warning=FALSE}

@@

# analysis dataframe
or_may_sep_df <- or_ts_df %>% 
  # create day of the week variable
  mutate(day = as.factor(weekdays(date))) %>% 
  # set missing outcome values to 0 since missing indicates no ER or urgent care 
  # visits on that date (which is reasonable in sparsely populated counties)
  # mutate_each(funs(wo_miss = ifelse(is.na(.), 0, .)), n_obs:broken_arm_n) %>% 
  # create a binary smoke indicator variable based smk variables >5, >10, and >15 units 
  mutate(wrf_smk5 = ifelse(wrf_smk_pm >= 5, 1, 0), 
         wrf_smk10 = ifelse(wrf_smk_pm >= 10, 1, 0),
         wrf_smk15 = ifelse(wrf_smk_pm >= 15, 1, 0), 
         geo_smk5 = ifelse(geo_smk_pm >= 5, 1, 0),
         geo_smk10 = ifelse(geo_smk_pm >= 10, 1, 0),
         geo_smk15 = ifelse(geo_smk_pm >= 15, 1, 0),
         krig_smk5 = ifelse(krig_smk_pm >= 5, 1, 0),
         krig_smk10 = ifelse(krig_smk_pm >= 10, 1, 0),
         krig_smk15 = ifelse(krig_smk_pm >= 15, 1, 0),
         season = ifelse(date >= "2013-05-01" & date <= "2013-06-21", "spring", 
                  ifelse(date >= "2013-06-22" & date <= "2013-09-22", "summer",
                  ifelse(date >= "2013-09-23" & date <= "2013-12-21", "fall",
                                "other")))) %>% 
  # rescale pm variables to 10 units
  mutate_each(funs(unit10 = ./10), wrf_pm:krig_smk_pm) %>%
  select(-42, -43, -44)
  # select(1:2, 31:33, 36, 34:35, 38:44, 28:29, 30, 55, 58, 
  #        65, 64, 63) # subset and order variables used in analysis # 23 column
  # wa: 12 pm; 15 disease

@@

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('WRF-Chem Smoke', 'Geo-Weighted Smoke', 'Kriging Smoke')

@@

#glimpse(wash_jul_oct_df)
# finding columns to retain
#which(colnames(wash_jul_oct_df)=='geo_smk_pm_unit10')
#summary(wash_jul_oct_df)
# loop through outcomes
for(i in 6:18){

  # empty matrix for table
  point_estimates <- matrix(nrow = 3, ncol = 5, byrow = T)

@@

  # loop through pm estimation methods
  for(j in 41:43){

    var_name <- method_list[j-40]

@@

    # populate matrix
    row_n <- j-40
    # method
    point_estimates[row_n, 2] <- var_name
    # rate ratio
    point_estimates[row_n, 3] <- round(exp(model_df[2,2]), 3)
    # 95% lower bound
    point_estimates[row_n, 4] <- round(exp((model_df[2,2]) -
                                             1.96*(model_df[2,3])), 3)
    # 95% upper bound
    point_estimates[row_n, 5] <- round(exp((model_df[2,2]) +
                                             1.96*(model_df[2,3])), 3)

@@

  # combine previous values in dataframe that has all outcome/methods comparisons
  datalist[[i-5]] <- point_est_df

@@

kable(combined_point_est_df)

# ggplot of odds ratios, facet wrapped by outcomes -----
# convert variables from character to either numeric or factor
# factor preserves the order of the variable
combined_point_est_df$outcome <- factor(combined_point_est_df$outcome, 
                                        levels = unique(combined_point_est_df$outcome))

combined_point_est_df$pm_method <- factor(combined_point_est_df$pm_method, 
                                          levels = unique(combined_point_est_df$pm_method))

combined_point_est_df$risk_ratio <- as.numeric(combined_point_est_df$risk_ratio)
combined_point_est_df$lower95 <- as.numeric(combined_point_est_df$lower95)
combined_point_est_df$upper95 <- as.numeric(combined_point_est_df$upper95)

## ggplot
print_plot <- ggplot(combined_point_est_df, 
                     aes(x = pm_method, y = risk_ratio, colour = pm_method)) +
  geom_point() + #geom_text(vjust = 0, nudge_x = 0.3) +
  geom_errorbar(aes(ymin=lower95, ymax=upper95), width = 0.2) +
  facet_wrap(~outcome, nrow = 3) +
  geom_hline(yintercept = 1, linetype=2) +
  ggtitle('County-Level Time-Series: Association Between PM2.5 from \n Wildfire Smoke on Hospitalizations') +
  ylab('Risk Ratio for 10µg/m^3 Increase in PM2.5') +
  xlab('Time-Stratified Within July to October Fire Season') +
  scale_colour_discrete(name= "Smoke Method") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_text(angle = 90),
        # facet text size
        strip.text = element_text(size = 7),
        legend.text = element_text(size = 7))

print(print_plot)

ggsave("../plot_new/or_county_time_series.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

```

### 2017-11-03 r_scripts/analysis/or_dist_lag_model.Rmd

@@

@@

### Medication Distributed Lag

>>>>

  # dataframe for analysis creation
  # bind columns back together 
  df_analysis <- cbind(covariates_df, pm_estimates_df) %>%
    # remove missing pm values
    filter(!is.na(wrf_smk_pm_lag2_zip)) %>% 
    filter(!is.na(krig_pm_lag2_zip)) %>% 
    filter(!is.na(geo_wt_pm_lag2_zip)) %>%

@@

print(print_plot)
# save figure
ggsave("../plot_lag/fig_lag2.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

### 2017-11-03 r_scripts/analysis/or_result_zip_smoke.Rmd

```{r data import, include = F, eval = FALSE}

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../data_new/casecrossover/ER")

resp_casecross <- read_csv(paste(path, 
                                 "or_respiratory_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                                   "or_asthma_may_to_sep_time_strat_casecross_er.csv", 
                                   sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                                 "or_copd_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                                  "or_pneumonia_may_to_sep_time_strat_casecross_er.csv",
                                  sep="/"))
 # acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                         "or_acute_bronch_may_to_sep_time_strat_casecross_er.csv",
                                         sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                                       "or_arrhythmia_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_er.csv",
                                      sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_er.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                               "or_myocardial_infarc_may_to_sep_time_strat_casecross_er.csv", sep="/"))

# broken arm
broken_arm_casecross <- read_csv(paste(path, 
                                       "or_broken_arm_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))

@@

In these comparisons, we examine various methods of smoke/PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures. However, we found the filtered population is much larger than the population we chose in Washington, so we choose only the emegency room (hospital) visits. For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

@@

print(print_plot)
# save figure
ggsave("../plot_new/fig_only_er.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")
```

@@

  print(print_plot)
  # save figure
  ggsave("../plot_new/age_fig_three.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

  print(print_plot)
  # save figure
  ggsave("../plot_new/sex_fig_three.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

>>>>

```{r ER data import, include = F, eval = FALSE}

@@

```{r ER and urgent data import, include = F, eval = FALSE}

### Limit to ER visits and urgent care
## pos = 23 or pos = 20
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

  or_disease <- or_disease %>% 
    # filter the unknown gender
    filter(gender!= "U")

  # Create a permanent case-cross over dataset
  file_name <- paste('../data_new/casecrossover/ER_Urgent/or', i, 'may_to_sep_time_strat_casecross_both.csv', sep = "_")

num_person_both <- data.frame(rbind(var_list, N))
```

```{r urgent data import, include = F, eval = FALSE}

### Limit to ER visits and urgent care
## pos = 23 or pos = 20
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

  or_disease <- or_disease %>% 
    filter(pos=="20") %>%
    # filter the unknown gender
    filter(gender!= "U")

  # Create a permanent case-cross over dataset
  file_name <- paste('../data_new/casecrossover/Urgent/or', i, 'may_to_sep_time_strat_casecross_ur.csv', sep = "_")

num_person_ur <- data.frame(rbind(var_list, N))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../data_new/casecrossover/ER_Urgent")

resp_casecross <- read_csv(paste(path, 
                                 "or_respiratory_may_to_sep_time_strat_casecross_both.csv", 
                                 sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                                   "or_asthma_may_to_sep_time_strat_casecross_both.csv", 
                                   sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                                 "or_copd_may_to_sep_time_strat_casecross_both.csv", 
                                 sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                                  "or_pneumonia_may_to_sep_time_strat_casecross_both.csv",
                                  sep="/"))
 # acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                         "or_acute_bronch_may_to_sep_time_strat_casecross_both.csv",
                                         sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_both.csv",
                                sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                                       "or_arrhythmia_may_to_sep_time_strat_casecross_both.csv",
                                       sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_both.csv",
                                      sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_both.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_both.csv",
                                sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                               "or_myocardial_infarc_may_to_sep_time_strat_casecross_both.csv", sep="/"))

# broken arm
broken_arm_casecross <- read_csv(paste(path, 
                                       "or_broken_arm_may_to_sep_time_strat_casecross_both.csv",
                                       sep="/"))

@@

In these comparisons, we examine various methods of smoke/PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures. However, we found the filtered population is much larger than the population we chose in Washington, so we choose only the emegency room (hospital) visits and urgent care visits. For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

@@

print(print_plot)
# save figure
ggsave("../plot_new/fig_both_er_ur.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")
```

@@

  print(print_plot)
  # save figure
  ggsave("../plot_new/age_fig_three_both.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

  print(print_plot)
  # save figure
  ggsave("../plot_new/sex_fig_three_both.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")
